---
title: Tornados and excessive heats are the most harmful to population health while
  floods and hurricanes have the greatest economic consequences in the US
author: "Olga Larina"
date: "4/15/2019"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Synopsis

The U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database
tracks characteristics of major storms and weather events in the United States,
including when and where they occur, as well as estimates of any fatalities,
injuries, and property damage. This data analysis addresses the following questions:

1. Across the United States, which types of events are most harmful with respect
to population health?
2. Across the United States, which types of events have the greatest economic
consequences?

I cleaned data to get real types of events (according to categories from documentation), analized and summarized all fatalities, injuries, and property damages that have been recorded since 1993, because earlier only few types of events have been
recorded. As soon as I want to compare events, I have to to take years for which
there is more information. Most harmful with respect to population health are
tornados, excessive heats, floods, flash floods, lightnings and thunderstorm winds.
Floods, hurricanes, tornados and hails have the greatest economic consequences.

# Data Processing

I downloaded the data from the project task repository and filtered only variables
that are important for the project. Also, convert dates to Data type.
```{r cache=TRUE}
library(dplyr)
library(stringr)
library(stringdist)
library(ggplot2)
if (!file.exists("./data"))
    dir.create("./data")
if (!file.exists("./data/repdata_data_StormData.csv.bz2"))
{
    fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
    download.file(fileURL, "./data/repdata_data_StormData.csv.bz2",method = "curl")
}
data <- read.csv("./data/repdata_data_StormData.csv.bz2",na.strings = "")
subdata <- data[,c("BGN_DATE","EVTYPE","FATALITIES","INJURIES","PROPDMG", "PROPDMGEXP","CROPDMG","CROPDMGEXP")]
dates <- subdata$BGN_DATE
dates <- as.Date(as.character(dates),"%m/%d/%Y")
subdata$BGN_DATE <- dates
```
EVTYPE contains types of events.
```{r}
length(unique(subdata$EVTYPE))
```
Are there so many categories for all years? Let's see.
```{r}
library(dplyr)
subdata2 <- filter(subdata, BGN_DATE < as.Date("01/01/1993","%m/%d/%Y"))
unique(subdata2$EVTYPE)

subdata3 <- filter(subdata, BGN_DATE < as.Date("01/01/1994","%m/%d/%Y"))
length(unique(subdata3$EVTYPE))
```
So, for 1993 we have a lot of data. Let's take 1993 and futher.
```{r}
subdata <- filter(subdata, BGN_DATE >= as.Date("01/01/1993","%m/%d/%Y"))
```
From the documentation I know that there have to be 48 categories, let's create
variable for them.
```{r}
evTypes <- c("Astronomical Low Tide", "Avalanche", "Blizzard", "Coastal Flood", "Cold/Wind Chill","Debris Flow", "Dense Fog", "Dense Smoke", "Drought", "Dust Devil", "Dust Storm", "Excessive Heat", "Extreme Cold/Wind Chill", "Flash Flood", "Flood", "Frost/Freeze", "Funnel Cloud", "Freezing Fog", "Hail", "Heat", "Heavy Rain", "Heavy Snow", "High Surf", "High Wind", "Hurricane (Typhoon)", "Ice Storm", "Lake-Effect Snow", "Lakeshore Flood", "Lightning", "Marine Hail", "Marine High Wind", "Marine Strong Wind", "Marine Thunderstorm Wind", "Rip Current", "Seiche", "Sleet", "Storm Surge/Tide", "Strong Wind", "Thunderstorm Wind", "Tornado", "Tropical Depression", "Tropical Storm", "Tsunami", "Volcanic Ash", "Waterspout", "Wildfire", "Winter Storm", "Winter Weather")
evTypes <- tolower(evTypes)
```
So, I have a lot of typos in the data. I have to convert subdata$EVTYPE to this
48 categories. An easy step is to remove spaces and convert to lower case.
```{r}
library(dplyr)
library(stringr)
subdata <- mutate(subdata, EVTYPE = tolower(EVTYPE))
subdata <- mutate(subdata, EVTYPE = str_trim(EVTYPE))
```
I noticed, that there are abbreviations of words "wind" and "thunderstorm",
let's fix it.
```{r}
subdata$EVTYPE<-gsub("tstm","thunderstorm",subdata$EVTYPE,fixed = TRUE)
subdata$EVTYPE<-gsub("wnd","wind",subdata$EVTYPE,fixed = TRUE)
```
The main idea of data cleaning is to take each category from evTypes and try to
find it in subdata$EVTYPE.
```{r cache = TRUE}
grep("hurricane|typhoon",evTypes,value=T)
unique(grep("hurricane|typhoon",subdata$EVTYPE,value=T))
subdata$EVTYPE[grep("hurricane|typhoon",subdata$EVTYPE)] <- "hurricane (typhoon)"

grep("tide",evTypes,value=T)
unique(grep("tide",subdata$EVTYPE,value=T))
subdata$EVTYPE[grep("blow-out tides|blow-out tide",subdata$EVTYPE)] <- "astronomical low tide"
subdata$EVTYPE[grep("astronomical high tide|high wind and high tides|high tides|storm surge",subdata$EVTYPE)] <- "storm surge/tide"

grep("flood",evTypes,value=T)
unique(grep("flood",subdata$EVTYPE,value=T))
s<-subdata$EVTYPE
subdata$EVTYPE[grep("(flash.*flood(ing)?|flood(ing)?.*flash)",subdata$EVTYPE)] <- "flash flood"
subdata$EVTYPE[grep("(coastal.*flood(ing)?|flood(ing)?.*coastal|cstl.*flood(ing)?|flood(ing)?.*cstl)",subdata$EVTYPE)] <- "coastal flood"
subdata$EVTYPE[grep("(lake(shore)?.*flood(ing)?|flood(ing)?.*lake(shore)?)",subdata$EVTYPE)] <- "lakeshore flood"
floodPos <- grepl("flood",subdata$EVTYPE) & !grepl("flash",subdata$EVTYPE) &
        !grepl("coastal",subdata$EVTYPE) & !grepl("lakeshore",subdata$EVTYPE)
subdata$EVTYPE[floodPos] <- "flood"


unique(grep("extreme cold/wind chill",subdata$EVTYPE,value=T))
subdata$EVTYPE[grep("cold",subdata$EVTYPE)] <- "extreme cold/wind chill"

subdata$EVTYPE[grep("avalanche",subdata$EVTYPE)] <- "avalanche"
subdata$EVTYPE[grep("blizzard",subdata$EVTYPE)] <- "blizzard"

subdata$EVTYPE[grep("dense fog|fog|patchy dense fog",subdata$EVTYPE)] <- "dense fog"
subdata$EVTYPE[grep("freezing fog|ice fog",subdata$EVTYPE)] <- "freezing fog"

subdata$EVTYPE[grep("dense smoke|smoke",subdata$EVTYPE)] <- "dense smoke"
subdata$EVTYPE[grep("drought",subdata$EVTYPE)] <- "drought"

subdata$EVTYPE[grep("dust",subdata$EVTYPE)] <- "dust storm"

subdata$EVTYPE[grep("heat",subdata$EVTYPE)] <- "excessive heat"

subdata$EVTYPE[grep("frost|freeze",subdata$EVTYPE)] <- "frost/freeze"

subdata$EVTYPE[grep("cloud",subdata$EVTYPE)] <- "funnel cloud"

hailPos <- grepl("hail",subdata$EVTYPE) & !grepl("marine",subdata$EVTYPE)

subdata$EVTYPE[hailPos] <- "hail"

subdata$EVTYPE[grep("rain|wet",subdata$EVTYPE)] <- "heavy rain"

lakeSnowPos <- grepl("(lake.*snow|snow.*lake)",subdata$EVTYPE)
subdata$EVTYPE[lakeSnowPos] <- "lake-effect snow"
subdata$EVTYPE[grepl("snow",subdata$EVTYPE) & ! lakeSnowPos] <- "heavy snow"

subdata$EVTYPE[grep("surf",subdata$EVTYPE)] <- "high surf"

## wind
nonColdMarine <- !grepl("extreme cold/wind chill",subdata$EVTYPE)&!grepl("marine",subdata$EVTYPE)
wind <- unique(grep("wind",subdata$EVTYPE,value=T))

subdata$EVTYPE[grepl("(thunderstorm.*wind|wind.*thunderstorm)",subdata$EVTYPE)&nonColdMarine] <- "thunderstorm wind"
subdata$EVTYPE[grepl("(high.*wind|wind.*high)",subdata$EVTYPE)&nonColdMarine] <- "high wind"
subdata$EVTYPE[grepl("(strong.*wind|wind.*strong)",subdata$EVTYPE)&nonColdMarine] <- "strong wind"

subdata$EVTYPE[grepl("(chill.*wind|wind.*chill)",subdata$EVTYPE)] <- "extreme cold/wind chill"
unique(subdata$EVTYPE[grep("wind",subdata$EVTYPE)])
subdata$EVTYPE<-gsub("winds","wind",subdata$EVTYPE,fixed = TRUE)
subdata$EVTYPE<-gsub("thundertorm|tunderstorm|thundertsorm|thundeerstorm|thuderstorm|thunerstorm|thundestorm|thunderstrom|thunderestorm","thunderstorm",subdata$EVTYPE)

nonTH <- !grepl("thunderstorm",subdata$EVTYPE) & !grepl("high",subdata$EVTYPE)
subdata$EVTYPE[grepl("wind",subdata$EVTYPE) & nonColdMarine & nonTH] <- "strong wind"

subdata$EVTYPE[grep("ice",subdata$EVTYPE)] <- "ice storm"
subdata$EVTYPE[grep("lightning",subdata$EVTYPE)] <- "lightning"
subdata$EVTYPE[grep("rip current",subdata$EVTYPE)] <- "rip current"
subdata$EVTYPE[grep("sleet",subdata$EVTYPE)] <- "sleet"
s<-subdata$EVTYPE

subdata$EVTYPE[grep("tornado",subdata$EVTYPE)] <- "tornado"
subdata$EVTYPE[grep("tropical storm",subdata$EVTYPE)] <- "tropical storm"
subdata$EVTYPE[grep("volcanic",subdata$EVTYPE)] <- "volcanic ash"
subdata$EVTYPE[grep("wildfire",subdata$EVTYPE)] <- "wildfire"

subdata$EVTYPE[grep("winter storm",subdata$EVTYPE)] <- "winter storm"
subdata$EVTYPE[grep("winter weather|winter mix|wintery mix|winter weather mix|winter weather/mix",subdata$EVTYPE)] <- "winter weather"
subdata$EVTYPE[grep("warm|hot|record high temperatures|record temperatures",subdata$EVTYPE)] <- "heat"
subdata$EVTYPE[grep("dry",subdata$EVTYPE)] <- "drought"
subdata$EVTYPE[grep("summary|none|\\?",subdata$EVTYPE)] <- "others"

subdata$EVTYPE[grepl("thunderstor",subdata$EVTYPE) & !grepl("marine",subdata$EVTYPE)] <- "thunderstorm wind"
subdata$EVTYPE[grep("coastalstorm|coastal storm",subdata$EVTYPE)] <- "storm surge/tide"
subdata$EVTYPE[grep("cool",subdata$EVTYPE)] <- "extreme cold/wind chill"
```
So, now we have categories.
```{r}
length(unique(subdata$EVTYPE))
```
Let's try to parse it one more time and if it doesn't work - replace with empty string.
```{r cache = TRUE}
library(stringdist)
fixedType <- character(length(subdata$EVTYPE))
for(i in 1:length(subdata$EVTYPE))
{
    type <- subdata$EVTYPE[i]
    typeSplit <- strsplit(type,"/")
    for (j in 1:length(typeSplit[[1]]))
        num <- sum(which(amatch(evTypes,typeSplit[[1]][j],maxDist=2)==1),na.rm = TRUE)
        if (num > 0)
            fixedType[i] <- evTypes[num]
}
```

# Results
It's time to answer the questions. Let's take a look on fatalities and injuries.
```{r}
library(ggplot2)
populationHealth <- data.frame(Event.Type = fixedType, Fatalities = subdata$FATALITIES,
                               Injuries = subdata$INJURIES,stringsAsFactors = FALSE)
popH <- aggregate(.~Event.Type, data = populationHealth, sum)
popH$Event.Type[grep("^$",popH$Event.Type)] <- "others"

topF <- quantile(popH$Fatalities, probs = 0.9)
popHF <- filter(popH, popH$Fatalities> topF)
print("The most harmful events: Fatalities")
popHF

gg<- ggplot(popHF,aes(x=reorder(Event.Type,-Fatalities),y = Fatalities)) +
    geom_col() +
    labs(x="Type of events", title = "The most harmful events: Fatalities")
print(gg)

topI <- quantile(popH$Injuries, probs = 0.9)
popHI <- filter(popH, popH$Injuries> topI)
print("The most harmful events: Injuries")
popHI

gg2<- ggplot(popHI,aes(x=reorder(Event.Type,-Injuries),y = Injuries)) +
    geom_col() +
    labs(x="Type of events", title = "The most harmful events: Injuries")
print(gg2)
```

To calculate the cost of loss, I have to take a look on "PROPDMG", "PROPDMGEXP",
"CROPDMG", "CROPDMGEXP". The "CROPDMGEXP" is the exponent values for "CROPDMG"
(crop damage). In the sam way, "PROPDMG" and "PROPDMGEXP".
```{r prop, cache = TRUE}
rnum <- length(subdata$PROPDMGEXP)
prop <- numeric(rnum)
propExp <- as.character(subdata$PROPDMGEXP)
propDmg <- subdata$PROPDMG
for (i in 1:rnum)
{
    exp <- propExp[i]
    pr <- propDmg[i]
    if (!is.na(exp))
    {
        if (exp=="0")
            prop[i] <- pr*10^0
        if (exp=="1")
            prop[i] <- pr*10^1
        if (exp=="2")
            prop[i] <- pr*10^2
        if (exp=="3")
            prop[i] <- pr*10^3
        if (exp=="4")
            prop[i] <- pr*10^4
        if (exp=="5")
            prop[i] <- pr*10^5
        if (exp=="6")
            prop[i] <- pr*10^6
        if (exp=="7")
            prop[i] <- pr*10^7
        if (exp=="8")
            prop[i] <- pr*10^8
        if (exp == "B")
            prop[i] <- pr*10^9
        if (exp == "h" | exp == "H")
            prop[i] <- pr*10^2
        if (exp == "K")
            prop[i] <- pr*10^3
        if (exp == "m" | exp == "M")
            prop[i] <- pr*10^6
    }
}
```
In the same way, I count crop.
```{r crop, echo = FALSE, cache = TRUE}
crop <- numeric(rnum)
cropExp <- as.character(subdata$CROPDMGEXP)
cropDmg <- subdata$CROPDMG
for (i in 1:rnum)
{
    exp <- cropExp[i]
    cr <- cropDmg[i]
    if (!is.na(exp))
    {
        if (exp=="0")
            crop[i] <- cr*10^0
        if (exp=="2")
            crop[i] <- cr*10^2
        if (exp == "B")
            crop[i] <- cr*10^9
        if (exp == "K" | exp == "k")
            crop[i] <- cr*10^3
        if (exp == "m" | exp == "M")
            crop[i] <- cr*10^6
    }
}
```
Let's take a look on the damage.
```{r}
economCons <- data.frame(Event.Type = fixedType, Damage = prop + crop,stringsAsFactors = FALSE)
ec <- aggregate(Damage~Event.Type, data = economCons, sum)
ec$Event.Type[grep("^$",ec$Event.Type)] <- "others"

topD <- quantile(ec$Damage, probs = 0.9)
ecD <- filter(ec, ec$Damage > topD)
print("The most expensive events")
ecD

gg3<- ggplot(ecD,aes(x=reorder(Event.Type,-Damage),y = Damage)) +
geom_col() +
labs(x="Type of events", title = "The most expensive events")
print(gg3)
```